\section{RISC-V Platform}\label{risc-v-platform}

\subsection{Introduction}\label{introduction-2}

The RISC-V specification provides for multi-threading and multi-core
implementations. A core is defined as an implementation with its own
instruction fetch unit. A hardware thread, or \emph{hart}, is defined as
a processing engine with its own state. A core may contain multiple
hardware threads.

The RV12 implements a single core with a single hardware thread, based
on the User Instruction Set Architecture v2.2 and Supervisor Instruction
Set Architecture v1.9.1 specifications. See
\href{http://www.riscv.org}{www.riscv.org} for the
specifications\footnote{Full reference details of the specifications are
  documented in section 11}.

\subsection{Privilege Levels}\label{privilege-levels}

At any time, a hardware thread (\emph{hart}) is running at some
privilege level. The current privilege level is encoded in one or more
Control and Status Registers (CSRs). See section 7, ``Control \& Status
Registers'', for more information.

The RISC-V specification defines four privilege levels.

\begin{longtable}[]{@{}llll@{}}
\toprule
Level & Encoding & Name & Abbreviation\tabularnewline
\midrule
\endhead
0 & 00 & User/Application & U\tabularnewline
1 & 01 & Supervisor & S\tabularnewline
2 & 10 & Hypervisor & H\tabularnewline
3 & 11 & Machine & M\tabularnewline
\bottomrule
\end{longtable}

Table ‑: RISC-V Privilege Levels

Each level provides its own protection and isolation.

The highest privilege level is the Machine level. This is an inherent
trusted level and has access to, and can alter, the whole machine.

The lowest level is the User/Application level and is considered the
least trusted level. It is used to protect the rest of the system from
malicious applications.

Supervisor mode is used to provide isolation between an operating system
and the machine and user levels. Hypervisor mode is used to virtualize
operating systems.

The RV12 always implements Machine mode and optionally implements User
mode
an\protect\hypertarget{_Toc326677752}{}{\protect\hypertarget{_Ref326824571}{}{}}d
parts of the Supervisor Mode.
